Probabilities and propensities: conversion experiences in online grocery shopping

### 2017-07-26 log.rmd
starting a/b test
trying to realize themselves

>>>>

## Wed 26 Jul 2017 12:06:20 BST
- at the heart of the paper is the contrast between the association rules and the personal relevance model. The differences between them aren't as great as might seem to be ...   
- where I finished last time -- line 153 -- there is still a gap. I didn't get to say everything I wanted to about the logistic regression model;  I wanted to add some stuff about how the model includes all the different things
- starting adding the probability section and adding in a bit of the Popper argument  
- got to the point of trying to say what 'trying to realize themselves' might mean. I intend to connect this to both 'trying to construct good features.'

### 2017-07-26 reconstruction.rmd
trying to realize themselves

This paper is a philosophical reconstruction [@Dewey_2017] of an online grocery ordering system undergoing an upgrade. The reconstruction is empirical in certain respects. It departs from an ethnographic moment: being a member of the audience at an industry/academic conference where a data scientist was describing how recommendations were created for online grocery orders. Starting from this highly mediated and fragmentary encounter with a recommender system and one of its designers, the paper explores the problem the system addresses.  It draws on anthropological and sociological research concerning shopping and list-making to situate the recommender system amongst the social ordering practices of shopping and shopping lists. It makes use of archaeological approaches, in the sense developed by Michel Foucault [@Foucault_1972] in order to identify and map important functional statements and practices configuring the knowledge and order generated by such systems over time. It conducts several code-based experiments in order to reconstruct, using widely available code resources resources such as API (Application Programmer Interfaces) and software libraries for machine learning, some prototypical elements of the system in question.  In doing this, it draws  on the philosophy of science and media theory to frame an engagement with data, algorithms and recommendations around probabilities and their worldly enactments.

Like some recent work in science and technology studies,  anthropology and media studies [@Marres_2017; @Marcus_2014; @Bogost_2012], through this combination of approaches,  the paper allows its own reconstruction to be provisionally shaped by an experiential encounter with the object it describes. At times, this mode of empirical philosophical engagement with its objects may well seem overly concerned with the technical minutiae of the grocery ordering system, and lacking in  sociological rigour or empirical evidence. With some forbearance on the part of readers, however, the main philosophical argument of the paper should resonate sociologically. It concerns social order, or the consistency and regularity of relating and acting amid a world in where probabilities and calculations based on probabilities play a more pervasive role.

@@

@@

In his account of probabilities as physical propensities, Popper emphasises the specific configurations established in laboratory experiments.

> experiments. ... by creating, at will, artificial conditions that either exclude, or reduce to zero, all the interfering and disturbing propensities 23

@@

For instance, the predictions of the recommender system themselves are the subject of probabilistic treatment. Patel described the experimental deployment of  the relevancy model experimentally in a random A/B controlled trial on the Tesco website. All customers were  allocated to one of four categories as shown in the table. In the  A/B trial,  an identical shopping list, or even the same person, might receive different recommendations.   The randomised application of the model, drawing on protocols for randomised clinical trials first developed in the 1960s, is widely used in social media platforms and hence in the implementation and observation of the effects of recommender systems.[^238.1] Random allocation of customers to the four categories (Test A, Test B, Control A, Control B) adds a layer of probability to the recommender system in the name of statistical validation of the effects or the 'uplift' of the model.[^122] Ironically,  the effects of a predictive model cannot be known in advance. They can only be observed experimentally.

@@

> What may happen in the future .. is, to some extent, open. There are many possibilities trying to realize themselves, but few of them have a very high propensity, given the initial conditions. 22

>>>>

This paper philosophically reconstructs [@Dewey_2017] an online grocery ordering system undergoing an upgrade. The reconstruction is empirical in certain respects. It departs from an ethnographic moment: being a member of the audience at an industry/academic conference where a data scientist was describing how recommendations were created for online grocery orders. Starting from this highly mediated and fragmentary encounter with a recommender system and one of its designers, the paper explores the problem the system addresses.  It draws on anthropological and sociological research concerning shopping and list-making to situate the recommender system amongst the social ordering practices of shopping and shopping lists. It makes use of archaeological approaches, in the sense developed by Michel Foucault [@Foucault_1972] in order to identify and map important functional statements and practices configuring the knowledge and order generated by such systems over time. It conducts several small-scale code-based experiments in order to reconstruct, using widely available code resources resources such as API (Application Programmer Interfaces) and software libraries for machine learning, some prototypical elements of the system in question.  In doing this, it draws  on the philosophy of science and media theory to frame an engagement with data, algorithms and recommendations in terms of probabilities and their worldly enactments.

Like some recent work in science and technology studies,  anthropology and media studies [@Marres_2017; @Marcus_2014; @Bogost_2012], through this combination of approaches,  the paper allows its own reconstruction to be provisionally shaped by an experiential encounter with the object it describes. At times, this mode of empirical philosophical engagement with its objects may well seem overly concerned with the technical minutiae of the groceries and recommender systems, and lacking in  sociological rigour or empirical evidence. With some forbearance on the part of readers, however, the main philosophical argument of the paper should resonate sociologically. It concerns social order, or the consistency and regularity of relating and acting amid a world in where probabilities and calculations based on probabilities play a more pervasive role.

@@

Many other slight adjustments, tweaks and local modifications would probably occur around the logistic regression model. The model's recommendations themselves have propensities that need to be adjusted to the scale of values of grocery retail.  
[FEELS LIKE A JUMP HERE:  see notes 26 July 2017]

@@

In his account of probabilities as physical propensities, Popper emphasises the specific configurations established in laboratory experiments:

> experiments work ... by creating, at will, artificial conditions that either exclude, or reduce to zero, all the interfering and disturbing propensities [@Popper_1990,23].

Like many accounts of laboratory practice [@Latour_1097; @Rheinberger, @Knorr-Cetina; @Lynch - epistopics ref], Popper understands experiments as stabilizing the conditions so that phenomena can be saved from their own variability. Experiments play an important role in conversion events, or in the process of bring new predictive models to bear on shopping practice. A  crucial consideration for a predictive model in a recommender system will be how to experimentalize its own predictions so that they are not interfered with or disturbed by other propensities. This is obviously an inordinately difficult challenge, since, as we have seen, grocery shopping takes place at the intersection of normative and actual social orders, individual and group belongings, and logistically global economic processes.

@@

Nonetheless, experimentalization, the practice of creating conditions that reduce disturbing propensities, is very much part of the convernsion between models. For instance, the predictions of the recommender system themselves are the subject of experimental. Patel described the deployment of  the relevancy model experimentally in a random A/B controlled trial on the Tesco website. All customers were  allocated to one of four categories as shown in the table. In the  A/B trial,  an identical shopping list, or even the same person, might receive recommendations from different models.   The randomised application of  competing predictive models, draws on protocols for randomised clinical trials first developed in the 1960's, and is widely used in social media platforms and hence in the implementation and observation of the effects of recommender systems.[^238.1] Random allocation of customers to the four categories (Test A, Test B, Control A, Control B) adds a layer of probability to the recommender system in the name of statistical validation of the effects or the 'uplift' of the model.[^122] Ironically,  the effects of a predictive model cannot be known in advance. They can only be observed experimentally.

@@

Whatever disturbing factors or interferences the experimental trials of different predictive models reduce, a broader and ineluctable stream of propensities remains in operation. Popper's account of probabilities as propensities presents openness as concomitant to realization:

>What may happen in the future .. is, to some extent, open. There are many possibilities trying to realize themselves, but few of them have a very high propensity, given the initial conditions [@Popper_1990, 22].

There are two ways of reading this kind of statement. One is to see Popper as simply stating one of the consequencies of his physicalist understanding of probabilities as tendencies inherent in a situation. Another is to focus on the pivotal phrase 'possibilities trying to realize themselves' by asking what 'trying to realize' might mean. I would suggest that Patel's description of DunnHumby's work on the new personal relevance model is just such a 'trying to realize' in important ways.
